Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data

### 2021-01-20 paper/paper.Rmd
changes to intro, pre-processing and reply

\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour. Two-dimensional linear projections of points are sampled uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$) are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection. \textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the right-most plot.)}

@@

\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang]. These are designed to tackle large sample size, not dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches can be still be applied after the transformation, especially for large sample sizes.}

\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions. This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data. When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a unform distribution in the low-dimensional display. In general (for \textit{any} underlying data distribution) we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume. The observed projected distribution is representative of the original high-dimensional distribution in this sense.}

@@

Before applying the transformation in practice, it can be important to preprocess the data, as generally required with methods for the visualisation of high-dimensional data. In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions. This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation.

Scaling is commonly also combined with centering as a preprocessing step. This is less important here where we always center the projected data before applying our transformation (i.e. the projected mean defines the origin of our polar coordinate system in the plane).

Even relatively low numbers of dimensions, see e.g. examples with $p=10$, already lead to sizable piling and thus large distortions in the sage display. However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise. In this case it makes sense to first reduce the number of dimensions, for example using principal components analysis, and to adjust the $\gamma$ parameter if needed.

We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of data distributions. XXX specific examples? explicitly mention other transformations?

>>>>

\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour. Two-dimensional linear projections of data, generated by sampling uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$), are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection. \textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}

@@

\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays} [@chang]. \textcolor{blue}{These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.}

\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions. This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data. When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a unform distribution in the low-dimensional display. In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume. The observed projected distribution is representative of the original high-dimensional distribution in this sense.}

@@

\textcolor{red}{Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data. In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions. This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation. Scaling is commonly also combined with centering as a preprocessing step. This is less important in our implementation because we always center the projected data before applying our transformation (i.e. the projected mean defines the origin of our polar coordinate system in the plane). Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.}

\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}

\textcolor{red}{The third common pre-processing would involve a dimension reduction step. Even with a relatively low number of dimensions, e.g. example with $p=10$, we see noticeable piling. However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise. In this case it is advisable to first reduce the number of dimensions, for example using principal components analysis. Other dimension reduction techniques could be used. The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two. }

### 2021-01-20 paper/paper.tex
changes to intro, pre-processing and reply

\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.}
Figure \ref{fig:density} illustrates this crowding problem, as would be
experienced when using a tour. Two-dimensional linear projections of
points are sampled uniformly within \(p\)-dimensional hyperspheres
(\(p=3, 10, 100\)) are displayed as hexbin plots. Color indicates log
count of the bin, with yellow mapping the highest counts. As \(p\)
increases the density concentrates in the center of the projection.
\textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the right-most plot.)}

@@

\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang]. These are designed to tackle large sample size, not dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches can be still be applied after the transformation, especially for large sample sizes.}

\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions. This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data. When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a unform distribution in the low-dimensional display. In general (for \textit{any} underlying data distribution) we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume. The observed projected distribution is representative of the original high-dimensional distribution in this sense.}

@@

Before applying the transformation in practice, it can be important to
preprocess the data, as generally required with methods for the
visualisation of high-dimensional data. In particular, when working with
projections that combine multiple variables it is important to ensure a
common scale across the dimensions. This is especially important with
the sage display, where the overall scale (set by the maximum radius
\(R\)) determines the effect of the transformation.

Scaling is commonly also combined with centering as a preprocessing
step. This is less important here where we always center the projected
data before applying our transformation (i.e.~the projected mean defines
the origin of our polar coordinate system in the plane).

Even relatively low numbers of dimensions, see e.g.~examples with
\(p=10\), already lead to sizable piling and thus large distortions in
the sage display. However, we are often confronted with much higher
dimensional datasets in which many of the variables are primarily adding
noise. In this case it makes sense to first reduce the number of
dimensions, for example using principal components analysis, and to
adjust the \(\gamma\) parameter if needed.

We may also wish to apply non-linear transformations before visualising
the data or combining variables in a projection. One example would be to
use a log transformation in the case of a skewed distribution. These
types of transformations will generally change the distribution across
the high-dimensional volume (and in particular also the position of the
mean), and can thus make the sage display more useful for a wide range
of data distributions. XXX specific examples? explicitly mention other
transformations?

>>>>

\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.}
Figure \ref{fig:density} illustrates this crowding problem, as would be
experienced when using a tour. Two-dimensional linear projections of
data, generated by sampling uniformly within \(p\)-dimensional
hyperspheres (\(p=3, 10, 100\)), are displayed as hexbin plots. Color
indicates log count of the bin, with yellow mapping the highest counts.
As \(p\) increases the density concentrates in the center of the
projection.
\textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}

@@

\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays}
\citep{chang}.
\textcolor{blue}{These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.}

\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions. This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data. When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a unform distribution in the low-dimensional display. In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume. The observed projected distribution is representative of the original high-dimensional distribution in this sense.}

@@

\textcolor{red}{Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data. In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions. This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation. Scaling is commonly also combined with centering as a preprocessing step. This is less important in our implementation because we always center the projected data before applying our transformation (i.e. the projected mean defines the origin of our polar coordinate system in the plane). Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.}

\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}

\textcolor{red}{The third common pre-processing would involve a dimension reduction step. Even with a relatively low number of dimensions, e.g. example with $p=10$, we see noticeable piling. However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise. In this case it is advisable to first reduce the number of dimensions, for example using principal components analysis. Other dimension reduction techniques could be used. The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two. }
